Reasoning about a Capability Machine with Local Capabilities

### 2017-05-24 popl-2018/paper.tex
Added title and started changing syntax figure.
Updated syntax figure to still be very compact and fit in new format (readability questionable).
Figure with excerpt op sem updated to fit in column.

%% Title information
\title[Short Title]{Full Title}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Subtitle}                     %% \subtitle is optional
\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
\thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

@@

\begin{figure}
  \begin{align*}
    \addr   &\in& \Addrs &\isdef \nats\\
    w &\in&\Words &\isdef \ints + \Caps \\
    \perm   &\in& \Perms &::= \noperm \mid \readonly\mid \readwrite\mid \readwritel\mid \exec\mid \entry\mid \rwx\mid \rwlx\\
    \gl&\in&\Globals & ::= \glob \mid \local \\
     && \Caps  &::= \left\{
                       \begin{multlined}
                         ((\perm,\gl),\start,\addrend,\addr)\mid\\
                         \start,\addr\in\Addrs,\addrend \in
                         \Addrs\cup \{\infty\}
                       \end{multlined} \right\}\\
    r       &\in& \RegName&\bnfdef \pcreg\mid r_0\mid r_1\mid\ldots\\
    \reg &\in& \Regs  &\isdef \RegName \rightarrow \Words\\
    m&\in& \Heaps &\isdef \Addrs \rightarrow \Words \\
    \Phi    &\in& \ExecConfs  &\isdef \Regs \times \Heaps \\
    &&\Confs &::= \ExecConfs + \{\failed \} + \{\halted\} \times \Heaps \\
    \ms     &\in& \MemSegments &::= \Addrs \parfun \Words
  \end{align*}
  \begin{equation*}
  \begin{array}{rcl}
    \rv    &\in& \ints + \RegName \\
    i      &::=& 
                 \jmp{\lv} \mid 
                 \jnz{\lv}{\rv} \mid
                 \move{\lv}{\rv} \mid 
                 \load{\lv}{\hv} \mid 
                 \store{\hv}{\rv} \mid  \\
           &   & \plus{\lv}{\rv}{\rv} \mid 
                 \minus{\lv}{\rv}{\rv} \mid 
                 \lea{\lv}{\rv} \mid 
                 \restricttwo{\lv}{\rv} \mid \\
           &   & \subseg{\lv}{\rv}{\rv} \mid  
                 \isptr{\lv}{\rv} \mid 
                 \getp{\lv}{\lv} \mid 
                 \getl{\lv}{\lv} \mid\\ 
           &   & \getb{\lv}{\lv} \mid
                 \gete{\lv}{\lv} \mid
                 \geta{\lv}{\lv} \mid 
                 \fail \mid
                 \halt 
  \end{array}
\end{equation*}
  \caption{The syntax of our capability machine assembly language.}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure*}
  \centering
  \begin{equation*}
    \Phi  \rightarrow
    \begin{cases}
      \sem{\decode(n)}(\Phi) & \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$ and $\start \leq \addr \leq \addrend$ and $\perm \in \{ \exec,\rwx, \rwlx \}$ and $\memheap(\addr) = n$ }}\\
      \failed                                 & \text{\footnotesize{otherwise}}
    \end{cases}
  \end{equation*}\vspace{-.3cm}%
  \begin{align*}
    \stdUpdatePc{\Phi} &=
                         \begin{cases}
                           \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
                             \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$ and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}}\\
                             \failed & \text{\footnotesize{otherwise}}
                         \end{cases}
  \end{align*}
  \begin{tabular}{|c|p{3.5cm}|p{10.7cm}|}
    \hline
    $i$&$\sem{i}(\Phi)$&Conditions\\
    \hline 
    $\fail$&$\failed$&\\
    \hline
    $\halt$&$(\halted,\memheap)$&\\
    \hline
    $\move{\refreg{r_1}}{r_2}$& $\stdUpdatePc{}(\updateReg{r_1}{w})$&$w = \memreg(r_2)$\\
    \hline
    $\load{\refreg{r_1}}{\refheap{r_2}}$&$\stdUpdatePc{\updateReg{r_1}{w}}$&$\memreg(r_2) = \stdcap{}$ \footnotesize{ and }$w = \memheap(\addr)$\footnotesize{ and }$\start \leq \addr \leq \addrend$\newline \footnotesize{ and }$\perm \in \{ \rwx, \rwlx, \exec, \readwrite, \readwritel, \readonly \}$ \\
    \hline
    $\restricttwo{\refreg{r_1}}{r_2}$&$\stdUpdatePc{}(\updateReg{r_1}{w})$  & $\memreg(r_2) = \stdcap$\footnotesize{ and }$(\perm',g') = \decodePermPair{\memreg(r_2)}$
                                                      \footnotesize{ and }$(\perm',g') \sqsubseteq (\perm,g)$\footnotesize{ and } $w =((\perm',g'),\start,\addrend,\addr)$\\
    \hline
    $\geta{\refreg{r_1}}{\refreg{r_2}}$ & $\stdUpdatePc{\updateReg{r_1}{\addr}}$ &
                                                $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$\\
    \hline
    $\jmp{\lv}$&$\updateReg{\pcreg}{\var{newPc}}$& \footnotesize{if }$\memreg(r) = ((\entry,\gl),\start,\addrend,\addr)$\footnotesize{, then} $\var{newPc} = ((\exec,\gl),\start,\addrend,\addr)$\newline \footnotesize{otherwise }$\var{newPc} = \memreg(r)$\\
    \hline
    $\store{\refheap{r_1}}{\refreg{r_2}}$&$\stdUpdatePc{\updateHeap{\addr}{\var{w}}}$&$\memreg(r_1) = \stdcap$\footnotesize{ and }$\perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}$ \footnotesize{ and } $\start \leq \addr \leq \addrend$\footnotesize{ and }$\var{w} = \memreg(r_2)$
                                                                                       \footnotesize{and if }$\var{w} = ((\_,\local),\_,\_,\_)$\footnotesize{, then  }$\perm \in \{\rwlx,\readwritel \}$\\
    \hline
    \multicolumn{3}{|c|}{$\cdots$}\\
    \hline
    \_&$\failed$&\footnotesize{otherwise}\\
    \hline
  \end{tabular}
  \caption{An excerpt from the operational semantics.}
  \label{fig:op-sem}
\end{figure*}

@@

\begin{comment}
  \begin{figure}
    \centering
    \begin{mathpar}
      \Phi \rightarrow
      \begin{cases}
        \sem{\decode(n)}(\Phi) & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\
          \;\;\text{\footnotesize{and $\start \leq \addr \leq \addrend$ and $\memheap(\addr) = n$}}\\
          \;\;\text{\footnotesize{and $\perm \in \{ \exec,\rwx, \rwlx \}$} }
        \end{array}\\
        \failed                & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \updatePcPerm{w} =
      \begin{cases}
        ((\exec,\gl),\start,\addrend,\addr) & \text{\footnotesize{if $w = ((\entry,\gl),\start,\addrend,\addr)$}}\\
        w & \text{\footnotesize{otherwise}}
      \end{cases}\\
      \and \stdUpdatePc{\Phi} =
      \begin{cases}
        \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\ 
          \;\;\text{\footnotesize{and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}}
        \end{array}\\
        \failed & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \inferrule{ } { \sem{\fail}(\Phi) = \failed } \and
      \inferrule{ } { \sem{\halt}(\Phi) = (\halted,\memheap) } \and
      \inferrule{ } { \sem{\move{\refreg{r_1}}{r_2}}(\Phi) =
        \stdUpdatePc{}(\updateReg{r_1}{\memreg(r_2)}) } \and
      \inferrule{ \memreg(r_2) = \stdcap \\
        \perm \in \{ \rwx, \rwlx, \exec, \readwrite,
        \readwritel, \readonly \}\\
        \start \leq \addr \leq \addrend } {
        \sem{\load{\refreg{r_1}}{\refheap{r_2}}}(\Phi) =
        \updateReg{r_1}{\memheap(\addr)}} \and
      \inferrule{ \memreg(r_2) = \stdcap[\permp]\\
        \permp' = \decodePermPair{\memreg(r_2)}\\
        \permp'\sqsubseteq \permp } {
        \sem{\restricttwo{\refreg{r_1}}{r_2}}(\Phi) = \\
        \stdUpdatePc{}(\updateReg{r_1}{(\permp',\start,\addrend,\addr)})
      } \and \inferrule{ \memreg(r_2) = ((\_,\_),\_,\_,\addr) } {
        \sem{\geta{\refreg{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateReg{r_1}{\addr}}} \and \inferrule{
        \var{newPc} = \updatePcPerm{\memreg(\lv)} } {
        \sem{\jmp{\lv}}(\Phi) = \updateReg{\pcreg}{\var{newPc}} } \and
      \inferrule{ \memreg(r_1) = \stdcap\\
        \start \leq \addr \leq \addrend\\
        \perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}\\
        \var{w} = \memreg(r_2) \\
        \var{w} = ((\_,\local),\_,\_,\_)\Rightarrow\perm \in
        \{\rwlx,\readwritel \} } {
        \sem{\store{\refheap{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateHeap{\addr}{\var{w}}} } \and
    \end{mathpar}
    \caption{An excerpt from the operational semantics.}
    \label{fig:op-sem}
  \end{figure}
\end{comment}
\figurename~\ref{fig:op-sem} shows an excerpt of the operational semantics for a
few representative instructions. Essentially, a configuration $\Phi$ either
decodes and executes the instruction at $\memreg(\pcreg)$ if it is executable
and its address is in the valid range or otherwise fails. The table in the
figure shows for instructions $i$ the result of executing them in configuration
$\Phi$. $\fail$ and $\halt$ obviously fail and halt respectively. $\move{}{}$
simply modifies the register file as requested and updates the $\pcreg$ to the
next instruction using the meta-function $\stdUpdatePc{}$.

@@

%\dominique{``end up being defined recursively'' does not sound very insightful}
A region defines what its part of memory can legally look like. However,
validity of the memory's contents may itself depend on the state of other parts
of memory. To accommodate this, worlds need to be defined recursively: Worlds
describe memory, which can contain capabilities, whose safety is
world-dependent. Recursive worlds are a common requirement in Kripke models and
the following lemma uses the method of
\citet{Birkedal:2011:SKM:1926385.1926401,Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi &\; : \; \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions & = \{\revoked\} \uplus \\
                 &\;\quad \begin{multlined}
                     \{\temp\} \times \States \times \Rels \times \\
                     (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))
                   \end{multlined} \uplus \\
                 &\;\quad \begin{multlined}
                     \{\perma\} \times \States \times \Rels \times \\
                     (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
                   \end{multlined}
    \end{align*}
  and for $W, W' \in \Wor$
  \begin{align*}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{align*}
\end{theorem}
We use the $\Regions$ from Theorem~\ref{thm:world-existence} to define
the worlds:
\[
  \Worlds = \nats \finparfun \Regions
\]

@@

\subsubsection{World satisfaction}
A memory satisfies a world, written $\memSat{\ms}{W}$, if it can be partitioned
into disjoint parts such that each part is accepted by an active (permanent or
temporary) region.
\begin{align*}
  &\memSat{\ms}{W}
    \text{ iff }\\
  &\quad \left\{\begin{aligned}
        &\exists P : \activeReg{W} \rightarrow \HeapSegments \ldotp \hs = \biguplus_{r\in\activeReg{W}}P(r) \text{\footnotesize{ and }} \\[-.2cm]
        &\quad \forall r \in \activeReg{W} \ldotp \exists H,s \ldotp\\
        &\qquad W(r) = (\_,s,\_,\_,H) \text{\footnotesize{ and }} \npair[n]{P(r)} \in H(s)(\xi^{-1}(W))\\
      \end{aligned}\right.
\end{align*}

\subsection{Logical relation}
\afterpage{
\begin{figure}[htbp]
  \centering
%  \begin{withmathindent}{0cm}
    \begin{align*}
      & \observations \; : \;  \Worlds \nefun \UPred{\Regs \times \HeapSegments} \\
      & \observations (W) \defeq \\
      & \;\left\{ \npair{(\reg,\ms)} \; \middle| \;
        \begin{aligned}
          & \forall \ms_f, \heap', i \leq n \ldotp \\
          & (\reg,\ms \uplus \ms_f) \step[i] (\halted,\heap')  \\
          & \;\; \Rightarrow
          \begin{aligned}[t]
            & \exists W' \futurestr W, \ms_r, \ms' \ldotp\\
            & \;\; \heap' = \ms' \uplus \ms_r \uplus \ms_f \text{\footnotesize{ and }} \\
            & \;\; \heapSat[\hs']{n-i}{W'}
          \end{aligned}
        \end{aligned}\right\}\\
      & \stdrr \; : \; \Worlds \monwknefun \UPred{\Regs} \\
      & \stdrr(W) \defeq \left\{ \npair{\reg} \; \middle| \;
          \forall r \in \RegName \setminus \{\pcreg\} \ldotp \npair{\reg(r)}
          \in \stdvr(W) \right\} \\
      & \\[-.2cm]
      & \stder \; : \; \Worlds \nefun \UPred{\Words} \\
      & \stder(W) \defeq \left\{ \npair{\pc} \; \middle| \;
        \begin{multlined}
          \forall n' \leq n, \npair[n']{\reg} \in \stdrr(W), \heapSat[\hs]{n'}{W} \ldotp \\
          \npair[n']{(\reg\update{\pcreg}{\pc},\hs)} \in \observations(W)
        \end{multlined} \right\} \\
      & \\
      &\stdvr \; : \; \Worlds \monwknefun \UPred{\Words} \\
      &\stdvr(W)\defeq  \\
      &\;\;\begin{aligned}[t] & \{ \npair{i} \mid i \in \ints \}
        \union \\
        & \{ \npair{\stdcap[(\noperm,\gl)] } \}
        \union \\
        & \left\{
          \begin{multlined}
            \npair{\stdcap[(\readonly,\gl)] } \mid \\
            \npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
          \end{multlined} \right\}
        \union \\
        & \left\{
          \begin{aligned}
            &\npair{\stdcap[(\readwrite,\gl)] } \mid \\
             &\quad\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }} \\
             &\quad\npair{(\start,\addrend)} \in
            \writeCond{}(\iota^\nwl,\gl)(W)
          \end{aligned} \right\} \union \\
        & \left\{
          \begin{aligned}
            &\npair{\stdcap[(\readwritel,\gl)] } \mid \\
             &\quad\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }} \\
             &\quad\npair{(\start,\addrend)} \in
            \writeCond{}(\iota^\pwl,\gl)(W)
          \end{aligned} \right\}
        \union \\
        & \left\{
          \begin{aligned}
            &\npair{\stdcap[(\exec,\gl)]} \mid \\
             &\quad\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }} \\
             &\quad\npair{(\{\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
          \end{aligned} \right\}
        \union \\
        & \left\{
          \begin{multlined}
            \npair{\stdcap[(\entry,\gl)]} \mid \\

>>>>

%% Title information
\title{Reasoning about a Capability Machine with Local Capabilities}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Provably Safe Stack and Return Pointer Management (without OS Support)}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
%\thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

@@

\begin{figure}
\[
  \begin{array}{rlrcl p{2cm}rlrcl}
    \addr   &\in& \Addrs &\isdef& \nats & & r &\in& \RegName&\bnfdef& \pcreg\mid r_0\mid r_1\mid\ldots\\
    w &\in&\Words &\isdef& \ints + \Caps & & \reg &\in& \Regs  &\isdef& \RegName \rightarrow \Words\\
    \perm   &\in& \Perms &::= & \noperm \mid \readonly\mid \readwrite\mid \readwritel\mid  & & m&\in& \Heaps &\isdef& \Addrs \rightarrow \Words \\
    & & & &\exec\mid \entry\mid \rwx\mid \rwlx & & \Phi &\in& \ExecConfs  &\isdef& \Regs \times \Heaps \\
    \gl&\in&\Globals & ::= &\glob \mid \local & & \ms &\in& \MemSegments &::= & \Addrs \parfun \Words \\
    &&\Confs &::=& \multicolumn{5}{l}{\ExecConfs + \{\failed \} + \{\halted\} \times \Heaps}\\
     && \Caps  &::= & \multicolumn{5}{l}{\left\{
                         ((\perm,\gl),\start,\addrend,\addr)\mid
                         \start,\addr\in\Addrs,\addrend \in
                         \Addrs\cup \{\infty\}\right\}}
  \end{array}
\]
  \begin{equation*}
  \begin{array}{rcl}
    \rv    &\in& \ints + \RegName \\
    i      &::=& 
                 \jmp{\lv} \mid 
                 \jnz{\lv}{\rv} \mid
                 \move{\lv}{\rv} \mid 
                 \load{\lv}{\hv} \mid 
                 \store{\hv}{\rv} \mid \plus{\lv}{\rv}{\rv} \mid 
                 \minus{\lv}{\rv}{\rv} \mid 
                 \lea{\lv}{\rv} \mid 
                 \restricttwo{\lv}{\rv} \mid \\
           &   & \subseg{\lv}{\rv}{\rv} \mid  
                 \isptr{\lv}{\rv} \mid 
                 \getp{\lv}{\lv} \mid 
                 \getl{\lv}{\lv} \mid \getb{\lv}{\lv} \mid
                 \gete{\lv}{\lv} \mid
                 \geta{\lv}{\lv} \mid 
                 \fail \mid
                 \halt 
  \end{array}
\end{equation*}
  \caption{The syntax of our capability machine assembly language.}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure*}
  \centering
  \begin{equation*}
    \Phi  \rightarrow
    \begin{cases}
      \sem{\decode(n)}(\Phi) & \arraycolsep=0pt
      \begin{array}[t]{l}
\text{if $\memreg(\pcreg) = \stdcap$ and $\start \leq \addr \leq \addrend$}\\ \text{\;\;and $\perm \in \{ \exec,\rwx, \rwlx \}$ and $\memheap(\addr) = n$ }
      \end{array}
\\
      \failed                                 & \text{otherwise}
    \end{cases}
  \end{equation*}
  \begin{align*}
    \stdUpdatePc{\Phi} &=
                         \begin{cases}
                           \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
                             \text{if $\memreg(\pcreg) = \stdcap$ and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}\\
                             \failed & \text{otherwise}
                         \end{cases}
  \end{align*}
  \begin{tabular}{|c|p{3.9cm}|p{8.3cm}|}
    \hline
    $i$&$\sem{i}(\Phi)$&Conditions\\
    \hline 
    $\fail$&$\failed$&\\
    \hline
    $\halt$&$(\halted,\memheap)$&\\
    \hline
    $\move{\refreg{r_1}}{r_2}$& $\stdUpdatePc{}(\updateReg{r_1}{w})$&$w = \memreg(r_2)$\\
    \hline
    $\load{\refreg{r_1}}{\refheap{r_2}}$&$\stdUpdatePc{\updateReg{r_1}{w}}$&$\memreg(r_2) = \stdcap{}$ and $w = \memheap(\addr)$ and$\start \leq \addr \leq \addrend$\newline and $\perm \in \{ \rwx, \rwlx, \exec, \readwrite, \readwritel, \readonly \}$ \\
    \hline
    $\restricttwo{\refreg{r_1}}{r_2}$&$\stdUpdatePc{}(\updateReg{r_1}{w})$  & $\memreg(r_2) = \stdcap$ and $(\perm',g') = \decodePermPair{\memreg(r_2)}$
                                                      and $(\perm',g') \sqsubseteq (\perm,g)$ and  $w =((\perm',g'),\start,\addrend,\addr)$\\
    \hline
    $\geta{\refreg{r_1}}{\refreg{r_2}}$ & $\stdUpdatePc{\updateReg{r_1}{\addr}}$ &
                                                $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$\\
    \hline
    $\jmp{\lv}$&$\updateReg{\pcreg}{\var{newPc}}$& if $\memreg(r) = ((\entry,\gl),\start,\addrend,\addr)$, then $\var{newPc} = ((\exec,\gl),\start,\addrend,\addr)$\newline otherwise $\var{newPc} = \memreg(r)$\\
    \hline
    $\store{\refheap{r_1}}{\refreg{r_2}}$&$\stdUpdatePc{\updateHeap{\addr}{\var{w}}}$&$\memreg(r_1) = \stdcap$ and $\perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}$  and $\start \leq \addr \leq \addrend$ and $\var{w} = \memreg(r_2)$
                                                                                       and if $\var{w} = ((\_,\local),\_,\_,\_)$, then $\perm \in \{\rwlx,\readwritel \}$\\
    \hline
    \multicolumn{3}{|c|}{$\cdots$}\\
    \hline
    \_&$\failed$&otherwise\\
    \hline
  \end{tabular}
  \caption{An excerpt from the operational semantics.}
  \label{fig:op-sem}
\end{figure*}

@@

\begin{comment}
  \begin{figure}
    \centering
    \begin{mathpar}
      \Phi \rightarrow
      \begin{cases}
        \sem{\decode(n)}(\Phi) & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{if $\memreg(\pcreg) = \stdcap$}\\
          \;\;\text{and $\start \leq \addr \leq \addrend$ and $\memheap(\addr) = n$}\\
          \;\;\text{and $\perm \in \{ \exec,\rwx, \rwlx \}$} 
        \end{array}\\
        \failed                & \text{otherwise}
      \end{cases}
      \and \updatePcPerm{w} =
      \begin{cases}
        ((\exec,\gl),\start,\addrend,\addr) & \text{\footnotesize{if $w = ((\entry,\gl),\start,\addrend,\addr)$}}\\
        w & \text{otherwise}
      \end{cases}\\
      \and \stdUpdatePc{\Phi} =
      \begin{cases}
        \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{if $\memreg(\pcreg) = \stdcap$}\\ 
          \;\;\text{and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}
        \end{array}\\
        \failed & \text{otherwise}
      \end{cases}
      \and \inferrule{ } { \sem{\fail}(\Phi) = \failed } \and
      \inferrule{ } { \sem{\halt}(\Phi) = (\halted,\memheap) } \and
      \inferrule{ } { \sem{\move{\refreg{r_1}}{r_2}}(\Phi) =
        \stdUpdatePc{}(\updateReg{r_1}{\memreg(r_2)}) } \and
      \inferrule{ \memreg(r_2) = \stdcap \\
        \perm \in \{ \rwx, \rwlx, \exec, \readwrite,
        \readwritel, \readonly \}\\
        \start \leq \addr \leq \addrend } {
        \sem{\load{\refreg{r_1}}{\refheap{r_2}}}(\Phi) =
        \updateReg{r_1}{\memheap(\addr)}} \and
      \inferrule{ \memreg(r_2) = \stdcap[\permp]\\
        \permp' = \decodePermPair{\memreg(r_2)}\\
        \permp'\sqsubseteq \permp } {
        \sem{\restricttwo{\refreg{r_1}}{r_2}}(\Phi) = \\
        \stdUpdatePc{}(\updateReg{r_1}{(\permp',\start,\addrend,\addr)})
      } \and \inferrule{ \memreg(r_2) = ((\_,\_),\_,\_,\addr) } {
        \sem{\geta{\refreg{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateReg{r_1}{\addr}}} \and \inferrule{
        \var{newPc} = \updatePcPerm{\memreg(\lv)} } {
        \sem{\jmp{\lv}}(\Phi) = \updateReg{\pcreg}{\var{newPc}} } \and
      \inferrule{ \memreg(r_1) = \stdcap\\
        \start \leq \addr \leq \addrend\\
        \perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}\\
        \var{w} = \memreg(r_2) \\
        \var{w} = ((\_,\local),\_,\_,\_)\Rightarrow\perm \in
        \{\rwlx,\readwritel \} } {
        \sem{\store{\refheap{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateHeap{\addr}{\var{w}}} } \and
    \end{mathpar}
    \caption{An excerpt from the operational semantics.}
    \label{fig:op-sem}
  \end{figure}
\end{comment}
\figurename~\ref{fig:op-sem} shows an excerpt of the operational semantics for a
few representative instructions. Essentially, a configuration $\Phi$ either
decodes and executes the instruction at $\memreg(\pcreg)$ if it is executable
and its address is in the valid range or otherwise fails. The table in the
figure shows for instructions $i$ the result of executing them in configuration
$\Phi$. $\fail$ and $\halt$ obviously fail and halt respectively. $\move{}{}$
simply modifies the register file as requested and updates the $\pcreg$ to the
next instruction using the meta-function $\stdUpdatePc{}$.

@@

%\dominique{``end up being defined recursively'' does not sound very insightful}
A region defines what its part of memory can legally look like. However,
validity of the memory's contents may itself depend on the state of other parts
of memory. To accommodate this, worlds need to be defined recursively: Worlds
describe memory, which can contain capabilities, whose safety is
world-dependent. Recursive worlds are a common requirement in Kripke models and
the following lemma uses the method of
\citet{Birkedal:2011:SKM:1926385.1926401,Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi \; : \; & \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions  = & \; \{\revoked\} \uplus \\
               & \; \{\temp\} \times \States \times \Rels \times (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))\uplus \\
               & \; \{\perma\} \times \States \times \Rels \times (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
    \end{align*}
  and for $W, W' \in \Wor$
  \begin{align*}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{align*}
\end{theorem}
We use the $\Regions$ from Theorem~\ref{thm:world-existence} to define
the worlds:
\[
  \Worlds = \nats \finparfun \Regions
\]

@@

\subsubsection{World satisfaction}
A memory satisfies a world, written $\memSat{\ms}{W}$, if it can be partitioned
into disjoint parts such that each part is accepted by an active (permanent or
temporary) region.
\begin{align*}
  &\memSat{\ms}{W}
    \text{ iff }\quad \left\{\begin{aligned}
        &\exists P : \activeReg{W} \rightarrow \HeapSegments \ldotp \hs = \biguplus_{r\in\activeReg{W}}P(r) \text{ and } \\[-.2cm]
        &\quad \forall r \in \activeReg{W} \ldotp \exists H,s \ldotp W(r) = (\_,s,\_,\_,H) \text{ and } \npair[n]{P(r)} \in H(s)(\xi^{-1}(W))\\
      \end{aligned}\right.
\end{align*}
